# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

First I read in the csv file and name it **Activity**.

```{r}
Activity <- read.csv("activity.csv",nrows=17568,
                     colClasses=c("integer","Date","integer"))
```


## What is the mean total number of steps taken per day?

Next I create a new dataframe of two columns: date and total steps taken 
that day.

```{r}
library(plyr)
DailySteps <- ddply(Activity, "date", summarize,sum(steps,na.rm=T))
colnames(DailySteps)  <- c("date","steps")
```

Next I create a histogram of the total steps taken each day and highlight the 
inordinate number of days no steps were reported

```{r fig.width=10}
library(ggplot2)
DailySteps$tooksteps <- factor(
        ifelse(DailySteps$steps==0,0,1))
hist <- ggplot(data=DailySteps,aes(x=steps,fill=tooksteps))
hist <- hist + geom_histogram(binwidth=2000)
hist <- hist + ggtitle("Daily Step Histogram") + 
        theme(plot.title = element_text(size = 20,face="bold"))
hist + scale_fill_discrete(breaks=c(0),labels=c("No Recorded Steps")) + 
        guides(fill=guide_legend(title=NULL)) + 
        scale_x_continuous(breaks=seq(0,25000,by=2000))
```

Here are the mean and median number of steps taken per day:

```{r}
mean(DailySteps$steps)
median(DailySteps$steps)
```


## What is the average daily activity pattern?

The following code accomplishes three tasks.<br/>
        1. Creates a new dataframe with the mean number of steps per interval <br/>
        2. Plots this timeseries data<br/>
        3. Returns the 5 minute interval with the most steps on average<br/>

```{r}
IntervalSteps <- ddply(Activity, "interval", summarize,mean(steps,na.rm=T))
colnames(IntervalSteps)  <- c("interval","mean_steps")
line <- ggplot(data=IntervalSteps, aes(x=interval,y=mean_steps))
line <- line + geom_line() + scale_x_continuous(breaks=seq(0,2400,by=200))
line + ggtitle("Mean Steps By Time of Day")
IntervalSteps$interval[which(IntervalSteps$mean_steps==max(IntervalSteps$mean_steps))]
```


## Imputing missing values

There are a number of missing step values however.  Let's figure out how many...

```{r}
sum(is.na(Activity$steps))
```

That's a lot of missing data! Let's impute the missing data with the mean of
there respective intervals.  This will take two steps:<br/>
1. Add a column of interval means to **Activity** using the *join* function and the **IntervalSteps** dataframe we already created. <br/>
2. Impute these interval means for rows that have NA steps<br/>

```{r}
Activity <- join(Activity,IntervalSteps,by="interval")
NARows <- is.na(Activity$steps)
Activity$steps[NARows] <- Activity$mean_steps[NARows]
```

With the imputation successfully accomplished.  We can now replot our daily
steps histogram as well as calculate the mean/median daily step values, to see
if imputation caused any drastic changes.

```{r}
DailySteps2 <- ddply(Activity, "date", summarize,sum(steps,na.rm=T))
colnames(DailySteps2)  <- c("date","steps")
hist <- ggplot(data=DailySteps2,aes(x=steps))
hist <- hist + geom_histogram(binwidth=2000,fill="blue",alpha=0.6)
hist <- hist + ggtitle("Daily Step Histogram (Imputed)") + 
        theme(plot.title = element_text(size = 20,face="bold"))
hist + scale_fill_discrete(breaks=c(0),labels=c("Zero Steps Recorded")) + 
        guides(fill=guide_legend(title=NULL)) + 
        scale_x_continuous(breaks=seq(0,25000,by=2000))
mean(DailySteps2$steps)
median(DailySteps2$steps)
```

Both the mean and median have grown considerably.


## Are there differences in activity patterns between weekdays and weekends?
Next I would like to explore the data and see if there is considerable 
difference between the user's weekend and weekday activity.  
To do this we will need to create a logical vector for whether an observation 
occured on the weekend. Let's try this: 

```{r}
Activity$weekend <- 
        factor(ifelse(weekdays(Activity$date) == "Saturday" 
               | weekdays(Activity$date)== "Sunday","Weekend","Weekday"))
```

That seems to have done the trick.  

Next we will create a new dataframe that calculates the mean number of steps 
taken split by both time interval and weekday/weekend.

Once we have this, we can graph the timeseries step data for both
weekends and weekdays and compare to see if any major disparities exist.

```{r}
IntervalSteps2 <- ddply(Activity, c("interval","weekend"), summarize,mean(steps,na.rm=T))
colnames(IntervalSteps2)  <- c("interval","weekend","mean_steps")
```

With this new dataframe, we can graph the timeseries step data for both
weekends and weekdays and compare them to see if any major disparities exist.
```{r}
ggplot(data=IntervalSteps2, aes(x=interval,y=mean_steps))+geom_line(color="blue")+facet_grid(weekend~.)+ggtitle("Weekend vs Weekday Mean Steps")+scale_x_continuous(breaks=seq(0,2400,by=200))
```

Looks like someone likes to sleep-in and stay up late on the weekends!